---
title: "Lab 4"
author: "Jeffrey Kinard"
date: "10/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(dplyr)
library(tidyr)
library(ggplot2)
```
# Part one: Cleaning data  
```{r}
library(readxl)
defense <- read_excel('/Users/jeff/Downloads/cyclonesFootball2019.xlsx', sheet='Defensive') %>% 
  data.frame()
offense <- read_excel('/Users/jeff/Downloads/cyclonesFootball2019.xlsx', sheet='Offensive') %>% 
  data.frame()
bio <- read_excel('/Users/jeff/Downloads/cyclonesFootball2019.xlsx', sheet='Biography') %>% 
  data.frame()
defClean <- defense %>% 
  mutate(across(c(1:2), as.factor)) %>% 
  mutate(across(where(is.character), as.numeric))
offClean <- offense %>% 
  mutate(across(c(1:2), as.factor)) %>% 
  mutate(across(where(is.character), as.numeric))
feet <- bio %>% 
  separate(Height, c("Feet","Inches"), sep = "-") %>% 
  select(Feet) %>% 
  mutate(across(c(1), as.numeric))
inches <- bio %>% 
  separate(Height, c("Feet","Inches"), sep = "-") %>% 
  select(Inches) %>% 
  mutate(across(c(1), as.numeric))
height <- (feet * 12) + inches
bio$Height <- height %>% unlist() %>% unname()
bio$Weight <- as.numeric(bio$Weight)
bioClean <- bio

offClean %>% head()
defClean %>% head()
bioClean %>% head()
```

# Part two: Tidying  
  
**1. Reformat the defClean data frame into a tidy format using pivot_longer. The type of statistic (Tackles_Solo, Tackles_ASST, etc) is added as a new key column named stat.**
```{r}
defCleanTidy <- defClean %>% pivot_longer(c(3:11), names_to="stat", values_to="stat_value")
defCleanTidy %>% head()
```
  
**2. Compare the distributions of the defensive statistics. What defensive skills are rare?**  
```{r}
defCleanTidy %>% group_by(stat) %>% summarise(sum=sum(stat_value))
```

**3. Did ISU have better defense against Iowa or Notre Dame? Answer this question by creating a scatterplot with x- and y-axes being the number of solo tackles (of each player). A large number of solo tackles is an indicator of good defense.**
```{r}
defCleanTidy %>% filter(stat=="Tackles_Solo" & Opponent_Opponent=="Iowa") %>% select("Name", "stat_value") %>% ggplot(aes(x=Name, y=stat_value)) + geom_point() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

defCleanTidy %>% filter(stat=="Tackles_Solo" & Opponent_Opponent=="Notre Dame") %>% select("Name", "stat_value") %>% ggplot(aes(x=Name, y=stat_value)) + geom_point() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
  
**4. Separate the Hometown column into two columns, namely the city and the state. Print the first few rows of these two columns. (Hint: look at the sep= argument of separate)**  
```{r}
bioClean %>% 
  separate(Hometown, c("City","State"), sep = ", ") %>% 
  select(c("State", "City")) %>% 
  head()
```

**5. How many players are there from each state?**
```{r}
bioClean %>% 
  separate(Hometown, c("City","State"), sep = ", ") %>% 
  select(c("State")) %>% 
  group_by(State) %>% 
  tally(name="num_players")
```
  
# Part three: Joining data frames  
  
**1. Find which states do the ISU defensive and offensive players come from. Create an appropriate summary.**
```{r}
defClean %>% 
  inner_join(bioClean, by="Name") %>% 
  separate(Hometown, c("City","State"), sep = ", ") %>% 
  select(c("Name", "State"))%>% 
  distinct() %>% 
  group_by(State) %>% 
  tally(name="count") %>% 
  ggplot(aes(x=State, y=count)) + geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r}
offClean %>% 
  inner_join(bioClean, by="Name") %>% 
  separate(Hometown, c("City","State"), sep = ", ") %>% 
  select(c("Name", "State"))%>% 
  distinct() %>% 
  group_by(State) %>% 
  tally(name="count") %>% 
  ggplot(aes(x=State, y=count)) + geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

**2. How does the performance of quarterback Brock Purdy relate to the performance of the ISU team?**
```{r}
brock <- offClean %>% 
  filter(Name=="Purdy, Brock") %>% 
  select(c(2:12)) %>% 
  mutate_all(~replace(., is.na(.), 0)) %>% 
  pivot_longer(c(2:11), names_to="stat", values_to="purdy_value")
team <- offClean %>% 
  select(c(2:12)) %>% 
  mutate_all(~replace(., is.na(.), 0)) %>% 
  pivot_longer(c(2:11), names_to="stat", values_to="team_value") %>% 
  group_by(Opponent_Opponent, stat) %>% 
  summarise(team_value=sum(team_value))

brock %>% inner_join(team, by=c("Opponent_Opponent", "stat")) %>% pivot_longer(c(3:4), names_to="player", values_to="value") %>% ggplot(aes(x=stat, y=value, fill=player)) + geom_bar(position="dodge", stat="identity") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + facet_wrap(Opponent_Opponent ~ .)
```

**3. Which ISU player(s) made big progress over the last year? Compare the 2019 and the 2018 data.**
```{r}

```